---
title: "Mayo Clinic PBC Trial Survival Analysis"
author: "Jiahe Tian"
date: "2026-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```


## Research Question

Among randomized participants in the Mayo Clinic PBC trial, was there evidence of a difference in survival between the D-penicillamine and placebo treatment groups?  


## Data Source

This analysis uses data from the Mayo Clinic randomized trial in primary biliary cirrhosis (PBC), available through the `survival` R package and the Rdatasets repository.  

The dataset includes baseline clinical and laboratory measurements, treatment assignment, follow-up time, and survival status. The randomized portion of the study compared D-penicillamine with placebo.  

For this analysis, we restrict the study population to randomized participants with known treatment assignments.  



## Setup

```{r}
# Install the required package if it is not already available
if (!requireNamespace("survival", quietly = TRUE)) {
  install.packages(
    "survival",
    repos = "https://cloud.r-project.org"
  )
}

library(survival)
```


## Load Data

```{r}
url <- "https://vincentarelbundock.github.io/Rdatasets/csv/survival/pbc.csv"

pbc <- read.csv(url)

head(pbc)
```

## Data Inspection

Before preparing the analysis dataset, we inspect the dimensions, variables, treatment assignments, and outcome information in the original PBC dataset.

```{r}
dim(pbc)

names(pbc)

table(pbc$trt, useNA = "ifany")

table(pbc$status)
```

## Outcome Definition

The outcome of interest is time to death. Follow-up time (`time`) is measured in days.  

Death (`status = 2`) is treated as the event of interest. Participants who were alive at the end of follow-up (`status = 0`) are censored. Participants who underwent liver transplantation (`status = 1`) are censored at the time of transplantation.  

Therefore, a binary event indicator is created as:  

- `event = 1`: death  
- `event = 0`: censored or liver transplantation  

## Data Preparation

The analysis is restricted to participants with a randomized treatment assignment. Treatment is relabeled for readability, and a binary death indicator is created for the survival analysis.

```{r}
pbc_randomized <- pbc[!is.na(pbc$trt), ]

pbc_randomized$treatment <- ifelse(
  pbc_randomized$trt == 1,
  "D-penicillamine",
  "Placebo"
)

pbc_randomized$event <- ifelse(
  pbc_randomized$status == 2,
  1,
  0
)
```


### Data Checks

```{r}
dim(pbc_randomized)

table(pbc_randomized$treatment)

# check derived outcome variable
table(
  Status = pbc_randomized$status,
  Event = pbc_randomized$event
)
```

```{r}
# check if there is any missing value
colSums(
  is.na(
    pbc_randomized[, c("time", "event", "treatment")]
  )
)
```



## Descriptive Summary

We first summarize the number of randomized participants and deaths by treatment group before comparing the survival distributions.

```{r}
table(pbc_randomized$treatment)

aggregate(
  event ~ treatment,
  data = pbc_randomized,
  FUN = sum
)
```


## Kaplan–Meier Survival Analysis

Kaplan–Meier survival curves were estimated separately for the D-penicillamine and placebo groups. Death was treated as the event of interest, while participants who underwent liver transplantation or remained alive at the end of follow-up were censored.

```{r}
surv_object <- Surv(
  time = pbc_randomized$time,
  event = pbc_randomized$event
)

km_fit <- survfit(
  surv_object ~ treatment,
  data = pbc_randomized
)

km_fit
```


```{r}
plot(
  km_fit,
  xlab = "Follow-up Time (Days)",
  ylab = "Survival Probability",
  main = "Kaplan-Meier Survival Curves by Treatment Group",
  conf.int = FALSE,
  mark.time = FALSE,
  col = c("steelblue", "darkorange")
)

legend(
  "topright",
  legend = c("D-penicillamine", "Placebo"),
  lty = 1,
  col = c("steelblue", "darkorange")
)
```

*AI usage: I used ChatGPT for guidance on customizing the R Kaplan–Meier plot to make its appearance comparable to the Python version, particularly for matching the plot colors.*


## Log-Rank Test

A log-rank test was used to formally compare the survival distributions between the D-penicillamine and placebo groups.

```{r}
logrank_result <- survdiff(
  surv_object ~ treatment,
  data = pbc_randomized
)

logrank_result

p_value <- 1 - pchisq(
  logrank_result$chisq,
  df = 1
)

round(p_value, 3)
```


## Interpretation

The Kaplan–Meier curves for the D-penicillamine and placebo groups were generally similar over the follow-up period. The curves crossed at several points, and neither treatment group showed a consistently higher survival probability than the other.  

Using a significance level of $\alpha$ = 0.05, the log-rank test did not show a statistically significant difference in the survival distributions between the two treatment groups (p = 0.75). Since the p-value was greater than 0.05, we failed to reject the null hypothesis of equal survival distributions.  

Overall, this unadjusted analysis did not provide evidence of a difference in survival between participants assigned to D-penicillamine and those assigned to placebo.  



## Python vs. R Comparison

For this analysis, R was faster for me to write and was also the implementation I felt more confident using. I have more experience using R for statistical analyses, including survival analysis, so the workflow using `Surv()`, `survfit()`, and `survdiff()` was relatively direct and familiar. In contrast, the Python implementation required me to set up an appropriate environment and install the `lifelines` package before performing the survival analysis. I also needed to manage the Python environment carefully so that the analysis could be reproduced.    

I have slightly more confidence in the R implementation because of my greater familiarity with R and the well-established `survival` package. However, the Python and R implementations produced essentially the same Kaplan–Meier curves and log-rank test result (p = 0.75). The agreement between the two independent implementations increases my confidence in the overall analysis and suggests that the results were not dependent on the choice of software.   



